### 2022-04-08 LaTeX/report.tex
comments on report and small fix in Isabelle theory

@@

\end{titlepage}
\restoregeometry
\pagebreak
~
\pagebreak

@@

\section{Introduction}
\subsection{Academic context}
This research work was carried out as part of my curriculum at the french \href{https://mines-nancy.univ-lorraine.fr}{École des Mines de Nancy}. All documents such as codes or source papers are available on a \href{https://github.com/VTrelat/Tarjan}{GitHub repository}.

\subsection{Formal methods}
Formal methods are a field of computer science related to mathematical logic and reasoning. The whole purpose of the discipline is to ensure by a logical proof that a given algorithm is not only correct on its domain of definition, but also to find -- or define -- that domain. Formal methods find applications in a variety of fields, both concrete, such as the railway industry or self-driving cars, and abstract, such as computational architecture.\\
Altough a formal proof lies first on paper, the real formalisation starts when proofs are mechanised in a proof assistant.

\subsection{Isabelle (HOL)}
\begin{quote}
    Isabelle is a generic proof assistant. It allows mathematical formulas to be expressed in a formal language and provides tools for proving those formulas in a logical calculus.
\end{quote}
\begin{flushright}
    \href{https://isabelle.in.tum.de/}{isabelle.in.tum.de}
\end{flushright}
Isabelle is a really powerful low-level proof assistant coming with a higher order logic (HOL) proving environment making the proofs easily readible and comprehensible without adding any abstract overlay. The term "assistant" designates the fact that Isabelle has numerous tools allowing various automations in the proofs such as a theorem seeker or an automatic solver.

@@

\subsection{Graphs and their representation}
A graph \GG is the data $(\mathcal{V}, \mathcal{E})$ where:
\begin{itemize}
    \item $\mathcal{V}$ is a set of vertices
    \item $\mathcal{E} \subseteq \{(x,y) \in V^2\}$ is a set of edges\footnote{Note the use of the couple $(x, y)$ and not the pair $\{x, y\}$ that makes the graph directed.}.
\end{itemize}
Vertices will often be called nodes and edges will be represented through adjency lists for each node.\\
Let us give an example. Let \GG = $(\mathcal{V}, \mathcal{E})$ be the graph represented on figure \ref{fig:graph}.
\begin{figure}[!h]
    \ctikzfig{graph}
    \caption{Representation of \GG\label{fig:graph}}
\end{figure}
Thus, \VV = $\{0,1,2,3,4,5,6,7\}$ and
\begin{flushleft}
    ~~~~~~~~\EE = $\{(0,0),(0,1),(0,2),(0,3),$\\
    $~~~~~~~~~~~~~~~(1,4),(1,7),$\\
    $~~~~~~~~~~~~~~~(3,0),(3,1),(3,2),(3,5),$\\
    $~~~~~~~~~~~~~~~(4,3),(4,6),$\\
    $~~~~~~~~~~~~~~~(5,6),$\\
    $~~~~~~~~~~~~~~~(6,3),$\\
    $~~~~~~~~~~~~~~~(7,6)\}$
\end{flushleft}

This representation being somewhat long, adjency lists can be used instead and therefore it gives:
\begin{center}
    \texttt{G.adjency = [[0,1,2,3],[4,7],[0,1,2,5],[3,6],[6],[3],[6]]}
\end{center}
Thus, for all $i \in \{0,\ldots,7\}$, \texttt{G.adjency[i]} is the list of nodes to which node $i$ is connected -- \textit{i.e.} there is an directed edge from node $i$ to every node of \texttt{G.adjency[i]}.
\pagebreak

\section{Formalisation}
\subsection{Strongly connected components}
\subsubsection{Directed graphs}\label{sec:directedgraphs}
\begin{definition}
    For two vertices $x$ and $y$ of \VV, the relation "has an edge to" is noted "$\Rightarrow$" such that
    \begin{equation*}
        (x, y) \in \mathcal{E} \Longleftrightarrow x \Rightarrow y
    \end{equation*}
    By extension, if there is a path from $x$ to $y$ with more than one edge, the same notation is kept for the sake of simplicity.
    The reflexive and transitive closure of the relation $\Rightarrow$ is noted $\Rightarrow^*$.
\end{definition}

@@

\subsubsection{Examples}
\begin{figure}[!h]
    \centering
    \begin{subfigure}[t]{.49\textwidth}
        \ctikzfig{example2}
        \subcaption{Strongly connected graph}
    \end{subfigure}
    \begin{subfigure}[t]{.49\textwidth}
        \ctikzfig{example3}
        \subcaption{Not strongly connected graph}
    \end{subfigure}
    \caption{Basic example of what is a small SCC}
\end{figure}
\begin{figure}[!h]
    \ctikzfig{exampleSCC}
    \caption{Example of a graph where each colored set of node is a maximal set of SCC\label{fig:exampleSCC}}
\end{figure}

\begin{figure}[!h]
    \ctikzfig{exampleSCCreduced}
    \caption{Reduced visualization of the graph represented if figure \ref{fig:exampleSCC}}
\end{figure}

\subsection{Order of traversal and backtracking edges}
\subsubsection{DFS and \textit{num} value}
Tarjan's SCC algorithm basically lies on a depth-first search. The figure \ref{fig:dfs} shows an example of a DFS traversal on a simple directed graph.
\begin{figure}[!h]
    \ctikzfig{dfs}
    \caption{Example of a DFS\label{fig:dfs}}
\end{figure}

@@

\subsubsection{Backtracking edges}
\begin{definition}
    Given a graph \GG and an order of traversal in this graph, \textit{i.e.} each node of \GG has a unique value $\texttt{num} \in [0,~|\mathcal{V}|]~$ and two nodes $u$ and $v$, there is a backtracking edge from $v$ to $u$ if:
    \begin{equation*}
        \left\{
            \begin{array}{l}
                \texttt{u.num} < \texttt{v.num}\\
                v~\Rightarrow~u,~\textit{i.e. }~(v, u) \in \mathcal{E}\\
            \end{array}
        \right.
    \end{equation*}
    In this case, the backtracking edge from $v$ to $u$ is represented by $v \hookrightarrow u$.\\      
\end{definition}

@@

A more formal definition would be the following:
\begin{definition}
    Let $u$ be a node.
    \begin{equation*}
        \texttt{u.lowlink} = \min \{\texttt{w.num}~|~\exists~v\in \mathcal{V},~u \Rightarrow v \hookrightarrow w\}
    \end{equation*}
\end{definition}

\subsubsection{Example}
Let \GG be the graph given in fig. \ref{fig:exlowlink}. The order of traversal of the graph is given by the value \textit{num} for each node of \GG. The \textit{lowlink} value is also displayed.

\begin{figure}[!h]
    \ctikzfig{lowlink}
    \caption{A DFS was performed through \GG from $x_1$ and next to each node is represented the couple of value (\textit{num}, \textit{lowlink})\label{fig:exlowlink}}
\end{figure}

Now, backtracking edges can be highlighted w.r.t. the order of traversal. In fig. \ref{fig:exlowlink2}, they are represented as red dashed arrows.\\
Knowing the backtracking edges, all \textit{lowlink} values can be computed\footnote{In fact, they are refreshed during the DFS.}.\\
Let us take $x_8$ as an example: its \textit{lowlink} is equal to 2, which actually means that $x_6$\footnote{Because $\texttt{x6.num} = 2$} is its anchor -- or attachment node -- in their SCC, namely the green one. Indeed, $x_8$ is alone in its equivalence class\footnote{for the relation $\Rightarrow$}, and from all nodes linked by one of the backtracking edges of $x_8$, $x_6$ has the minimum value \textit{num}. Likewise, $x_3$ is its own attachment node in the green SCC since it is the first node visited when performing the DFS.

@@

\pagebreak
\section{Tarjan's algorithm}
Tarjan's algorithm is an efficient on-the-fly SCC computing algorithm \cite{tarjanMerz}. It basically perfoms a DFS while updating the \textit{num} and \textit{lowlink} values. All nodes are stored in a stack during the traversal until a backtracking edge is found. In this case, the \textit{lowlinks} are computed and all nodes are unstacked and saved in a SCC until a node verifying the equality between its \textit{num} and its \textit{lowlink} -- which has to occur -- is found. Then, the DFS goes on. The whole process is written in the following algorithm \ref{alg:tarjan}.\\
It can be shown that every node and edge are visited only once so the algorithm can achieve a linear complexity, \textit{i.e.} $\mathcal{O}(|\mathcal{V}|+|\mathcal{E}|)$.

\subsection{Description of Tarjan's algorithm}
\begin{algorithm}[H]\label{alg:tarjan}
    \SetAlgoLined
    \KwData{A graph \GG = (\VV, \EE)}
    \KwResult{A partition \texttt{SCCs} of \VV where each element of \texttt{SCCs} is a SCC of \GG}
    Initialize an empty stack \texttt{R}\;
    Initialize an empty set \texttt{SCCs}\;
    Let $\texttt{num} := 0$\;
    \ForAll{$v \in \mathcal{V}$}{
        \If{\texttt{v.num} is unefined}{
            SCC(v)\;
        }
    }
    \SetKwProg{Function}{function}{}{}
    \Function{SCC: $v \in \mathcal{V} \rightarrow \texttt{None}$}{
        \texttt{v.num = num}\;
        \texttt{v.lowlink = num}\;
        increment \texttt{num}\;
        Push $v$ in \texttt{R}\;
        \texttt{v.onStack = true}\;
        \ForAll{$w \in \text{POST}(v)$}{
            \If{\texttt{w.num} is undefined}{
                \texttt{v.lowlink} = $\min(\texttt{v.lowlink, w.lowlink})$\;
            }
            \ElseIf{\texttt{w.onStack}}{
                \texttt{v.lowlink} = $\min(\texttt{v.lowlink, w.num})$\;
            }
        }
        \If{\texttt{v.lowlink = v.num}}{
            Initialize an empty set \texttt{currentSCC}\;
            \Repeat{$\texttt{v.num} \neq \texttt{w.num}$}{
                Let $w := \texttt{R.pop()}$\;
                \texttt{w.onStack = false}\;
                $\texttt{currentSCC} = \texttt{currentSCC} \cup \{w\}$\;
            }
            $\texttt{SCCs} = \texttt{SCCs} \cup \texttt{currentSCC}$\;
        }
    }
    \caption{Tarjan's algorithm}
\end{algorithm}
\pagebreak

\section{A sequential set-based algorithm}
\subsection{Formalisation}
\begin{definition}[SCC mapping]
    In the following algorithm, the SCCs are progressively tracked in a collection of disjoint sets through a map $\mathcal{S} : \mathcal{V} \longrightarrow \mathcal{P}(\mathcal{V})$, where $\mathcal{P}(\mathcal{V})$ is the powerset of \VV, s.t. the following invariant is maintained:
    \begin{equation}\label{invariant}
        \forall v, w \in \mathcal{V}, w\in \mathcal{S}(v) \Longleftrightarrow \mathcal{S}(v) = \mathcal{S}(w)
    \end{equation}
\end{definition}

@@

    \caption{Sequential set-based SCC algorithm}
\end{algorithm}

@@

\subsection{Informal proof}
Note that this proof is said informal only because it is not formally automated. Both logical and mathematical arguments developped below are absolutely relevant.

@@

\begin{proof}
    Let $x \in \mathcal{V}$ be the following node to be visited during the execution of the algorithm \ref{alg:seqsetbased}: $x$ is pushed in \texttt{R}. Let $y \in \textsc{Post}(x)$. There are two cases:\\
    \begin{itemize}
        \item y has not been visited yet, \textit{i.e.} $y \notin \textsc{Visited}$. Thus, a DFS-like traversal is performed from $y$, so $y$ is pushed in \texttt{R} and $\mathcal{S}(y) = \{y\}$ because $y$ is alone in its equivalence class for the moment since it has not been visited yet.\\
        Therefore, $\mathcal{S}(x) \cap \mathcal{S}(y) = \varnothing$.

        \item y has already been visited, \textit{i.e.} $y \in \textsc{Visited}$. Then, $y$ was already pushed in \texttt{R} before $x$. Let $(x_i)_{1\leq i \leq n}$ be the first nodes of the stack s.t. $x_0 = x$ and $x_n = y$.\\
        In order to avoid writing $\texttt{R} = [\ldots, y, \ldots, x]$, let us define $\widetilde{R}$ the stack containing the first $n$ nodes in \texttt{R}, s.t. $\widetilde{R} = [y, \ldots, x] = [x_n, \ldots, x_0]$.\\
        Let us consider the worst case, \textit{i.e.} when
        \begin{equation*}
            \forall~1 \leq i \leq n,~\mathcal{S}(x_i) = \{x_i\}
        \end{equation*}
        So, the while loop has to go down to $y$ because all partial SCCs are disjoint. As the length of the stack \texttt{R} is bounded by $|\mathcal{V}|$, the algorithm terminates.\\
        $x_0$ is first unstacked and both $\mathcal{S}(x_0)$ and $\mathcal{S}(\texttt{R.top()}) = \mathcal{S}(x_1)$ are then united. The current state of $\mathcal{S}$ and $\widetilde{R}$ is:
        \begin{equation*}
            \left\{
                \begin{array}{l}
                    \mathcal{S} = \{\{x_0, x_1\}, \{x_2\}, \ldots, \{x_n\}, \ldots\}\\
                    \widetilde{R} = [x_n, \ldots, x_1]\\
                \end{array}
            \right.
        \end{equation*}
        Then, $x_1$ is unstacked and $\mathcal{S}(x_1)$ and $\mathcal{S}(x_2)$ are then united, so that:
        \begin{equation*}
            \left\{
                    \begin{array}{l}
                        \mathcal{S} = \{\{x_0, x_1, x_2\}, \{x_3\}, \ldots, \{x_n\}, \ldots\}\\
                        \widetilde{R} = [x_n, \ldots, x_2]\\
                    \end{array}
                \right.
        \end{equation*}
        Finally (by induction), $\mathcal{S} = \{x_0, \ldots, x_n\}$ and $\widetilde{R} = [y]$, \textit{i.e.} $\mathcal{S}(x)=\mathcal{S}(y)$. It is important to notice that $x=x_0, x_1, \ldots, x_{n-1}$ are no longer in the stack, so this operation kept the invariant true.
    \end{itemize}
\end{proof}

@@

\begin{proof}
    The disjointness of all on-stack partial SCCs is given by lemma \ref{lemma:disjointness}. Nodes from $\textsc{Visited} \setminus \textsc{Eexplored}$ are in \texttt{R} because they are being processed. So, $\textsc{Live} \subseteq \texttt{R}$.\\
    By L.6-7 of algorithm \ref{alg:seqsetbased}, $\textsc{Visited} \subseteq \texttt{R}$.\\
    L.9-10 ensure that no explored node is pushed in \texttt{R}.\\
    L.24-25 keep the invariant by unstacking explored nodes from \texttt{R}, so $\texttt{R} \cap \textsc{Explored} = \varnothing$. Thus, $\texttt{R} = \textsc{Visited} \setminus \textsc{Explored} = \textsc{Live}$.
\end{proof}

\begin{corollary}\label{cor:cor1}
    \begin{equation*}
        \forall v \in \textsc{Live}, \exists!~r \in \texttt{R} \cap \mathcal{S}(v), \mathcal{S}(v) = \mathcal{S}(r)
    \end{equation*}
\end{corollary}
\begin{proof}
    Let $v \in \textsc{Live} = \displaystyle{\biguplus_{v \in \texttt{R}} \mathcal{S}(v)}$. $v$ is in a unique partial SCC $\mathscr{S} := \mathcal{S}(v)$. Because of lemma \ref{lemma:disjointness}, there cannot exist $x \neq y \in \texttt{R}$ s.t. $\mathcal{S}(x) = \mathcal{S}(y) = \mathscr{S}$. Thus, there exists a unique $x \in \texttt{R}$ s.t. $\mathcal{S}(x) = \mathscr{S}$ (and $x\in \texttt{R} \cap \mathscr{S}$).
\end{proof}

@@

\begin{proof}
    Holds because of corollary \ref{cor:cor1}.
\end{proof}

>>>>

% editorial comments in the text or in marginal notes
% 1st argument: initials of the person making the comment,
% 2nd argument: comment to insert
\long\def\ednote#1#2{\par\noindent\framebox{\begin{minipage}{0.99\linewidth}\linespread{.7}\footnotesize #1: #2\end{minipage}}\par}
\newcommand{\edmargin}[2]{\marginpar{\raggedright\linespread{.7}\tiny #1: #2}}

@@

\end{titlepage}
\restoregeometry
\pagebreak

@@

\section{Introduction}
\subsection{Academic context}
This research work was carried out as part of my curriculum at the french\edmargin{sm}{French} \href{https://mines-nancy.univ-lorraine.fr}{École des Mines de Nancy}. All documents such as codes or source papers are available on a \href{https://github.com/VTrelat/Tarjan}{GitHub repository}.

\subsection{Formal methods}
Formal methods are a field of computer science related to mathematical logic and reasoning. The whole purpose of the discipline is to ensure by a logical proof that a given algorithm is not only correct on its domain of definition, but also to find -- or define -- that domain. Formal methods find applications in a variety of fields, both concrete, such as the railway industry or self-driving cars, and abstract, such as computational architecture.\\
Although a formal proof lies first on paper, the real formalisation starts when proofs are mechanised in a proof assistant.

\ednote{sm}{%
  First, a stylistic remark: do not end lines in a \LaTeX\ document using \texttt{\textbackslash\textbackslash} (except in \texttt{tabular} environments or similar). Use a blank line to start a new paragraph: the style will define the appearance of a paragraph break.

  More importantly, formal methods should not be equated with theorem proving. They are really about giving precise, mathematical definitions to computer science concepts. Although the purpose of giving such definitions is to enable formal verification, many techniques besides theorem proving, such as model-based testing, run-time monitoring, model checking etc.\ are used.
}

\subsection{Isabelle (HOL)}
\begin{quote}
    Isabelle is a generic proof assistant. It allows mathematical formulas to be expressed in a formal language and provides tools for proving those formulas in a logical calculus.
\end{quote}
\begin{flushright}
    \href{https://isabelle.in.tum.de/}{isabelle.in.tum.de}
\end{flushright}
Isabelle is a really powerful low-level\edmargin{sm}{What does ``low-level'' mean?} proof assistant coming with a higher order logic (HOL) proving environment making the proofs easily readable and comprehensible without adding any abstract overlay.\edmargin{sm}{I don't understand what that means.} The term ``assistant'' designates the fact that Isabelle has numerous tools allowing various automations in the proofs such as a theorem seeker or an automatic solver.

\ednote{sm}{%
  What is a ``theorem seeker''? In fact, ``assistant'' refers to the fact that the machine checks the proof provided by the user, in contrast to automatic theorem proving where the machine finds the proof itself. The tools for automation are intended to help the user write the proof at a conveniently high level, without needing to work at the level of a logical calculus, for example.
}

@@

\ednote{sm}{%
  First, there is a break of notation (\NN vs.\ \texttt{N}). Also, the ``dot'' notation for records is not used in Isabelle. Finally, as you say, the presentation corresponds to an algorithm. I'd first introduce graphs mathematically and refer to data structures only with respect to a particular algorithm.
}

\subsection{Graphs and their representation}
A graph \GG is the data $(\mathcal{V}, \mathcal{E})$ where:
\begin{itemize}
    \item $\mathcal{V}$ is a set of vertices
    \item $\mathcal{E} \subseteq \{(x,y) \in V^2\}$ is a set of edges\footnote{Note the use of the couple $(x, y)$ and not the pair $\{x, y\}$ that makes the graph directed.}.\edmargin{sm}{couple $\rightarrow$ (ordered) pair, pair $\rightarrow$ set}
\end{itemize}
Vertices will often be called nodes and edges will be represented through adjency lists\edmargin{sm}{Again, that's an algorithmic concept.} for each node.\\
Let us give an example. Let \GG = $(\mathcal{V}, \mathcal{E})$ be the graph represented on figure \ref{fig:graph}.
\begin{figure}[!h]
    \ctikzfig{graph}
    \caption{Representation of \GG\label{fig:graph}}
\end{figure}
Thus, \VV = $\{0,1,2,3,4,5,6,7\}$ and
\begin{flushleft}\edmargin{sm}{Use numbers, not letters, in the figure so that the correspondence becomes clearer.}
    ~~~~~~~~\EE = $\{(0,0),(0,1),(0,2),(0,3),$\\
    $~~~~~~~~~~~~~~~(1,4),(1,7),$\\
    $~~~~~~~~~~~~~~~(3,0),(3,1),(3,2),(3,5),$\\
    $~~~~~~~~~~~~~~~(4,3),(4,6),$\\
    $~~~~~~~~~~~~~~~(5,6),$\\
    $~~~~~~~~~~~~~~~(6,3),$\\
    $~~~~~~~~~~~~~~~(7,6)\}$
\end{flushleft}

This representation being somewhat long, adjacency lists can be used instead and therefore it gives:
\begin{center}
    \texttt{G.adjacency = [[0,1,2,3],[4,7],[0,1,2,5],[3,6],[6],[3],[6]]}
\end{center}
Thus, for all $i \in \{0,\ldots,7\}$, \texttt{G.adjacency[i]} is the list of nodes to which node $i$ is connected -- \textit{i.e.} there is a directed edge from node $i$ to every node of \texttt{G.adjacency[i]}.
\pagebreak

\section{Formalisation}
\subsection{Strongly connected components}
\subsubsection{Directed graphs}\label{sec:directedgraphs}

\ednote{sm}{%
  I suggest moving this material to section 2: all of this is plain math. You can then show how these definitions are written in Isabelle.
}
\begin{definition}
    For two vertices $x$ and $y$ of \VV, the relation ``has an edge to'' is noted ``$\Rightarrow$'' such that
    \begin{equation*}
        (x, y) \in \mathcal{E} \quad\Longleftrightarrow\quad x \Rightarrow y
    \end{equation*}
    By extension, if there is a path from $x$ to $y$ with more than one edge, the same notation is kept for the sake of simplicity.\edmargin{sm}{The two sentences seem to contradict each other, and in order to avoid confusion I suggest writing systematically $\Rightarrow^*$ in order to denote reachability.}
    The reflexive and transitive closure of the relation $\Rightarrow$ is noted $\Rightarrow^*$.
\end{definition}

@@

\subsubsection{Examples}
\begin{figure}[!h]
    \centering
    \begin{subfigure}[t]{.49\textwidth}
        \ctikzfig{example2}
        \ednote{sm}{The notion of connectedness has not been introduced, and what's the relationship to SCCs?}
        \subcaption{Strongly connected graph}
    \end{subfigure}
    \begin{subfigure}[t]{.49\textwidth}
        \ctikzfig{example3}
        \subcaption{Not strongly connected graph}
    \end{subfigure}
    \caption{Basic example of what is a small SCC}
\end{figure}
\begin{figure}[!h]
    \ctikzfig{exampleSCC}
\ednote{sm}{%
  ``Set of SCC'' is confusing: an SCC is a set, and you are not talking about sets of sets here.
}
    \caption{Example of a graph where each colored set of node is a maximal set of SCC\label{fig:exampleSCC}}
\end{figure}

\begin{figure}[!h]
    \ctikzfig{exampleSCCreduced}
    \caption{Reduced visualization of the graph represented in figure \ref{fig:exampleSCC}}
\end{figure}

\subsection{Order of traversal and backtracking edges}
\subsubsection{DFS and \textit{num} value}
Tarjan's SCC algorithm\edmargin{sm}{not mentioned yet} basically lies on a depth-first search. The figure \ref{fig:dfs} shows an example of a DFS traversal on a simple directed graph.
\begin{figure}[!h]
    \ctikzfig{dfs}
    \caption{Example of a DFS\label{fig:dfs}}
\end{figure}

@@

\subsubsection{Backtracking edges}
\begin{definition}
    Given a graph \GG and an order of traversal in this graph, \textit{i.e.} each node of \GG has a unique value $\texttt{num} \in [0,~|\mathcal{V}|]~$ and two nodes $u$ and $v$, there is a backtracking edge from $v$ to $u$ if:
    \begin{equation*}
        \left\{
            \begin{array}{l}
                \texttt{u.num} < \texttt{v.num}\\
                v~\Rightarrow~u,~\textit{i.e. }~(v, u) \in \mathcal{E}\\
            \end{array}
        \right.
    \end{equation*}
    In this case, the backtracking edge from $v$ to $u$ is represented by $v \hookrightarrow u$.\edmargin{sm}{add example for the graph of Fig.5}\\      
\end{definition}

@@

A more formal definition would be\edmargin{sm}{is?} the following:
\begin{definition}
    Let $u$ be a node.
    \begin{equation*}
        \texttt{u.lowlink} = \min \{\texttt{w.num}~|~\exists~v\in \mathcal{V},~u \Rightarrow v \hookrightarrow w\}
    \end{equation*}
\end{definition}

\subsubsection{Example}
Let \GG be the graph given in fig. \ref{fig:exlowlink}.\edmargin{sm}{``Figure'' vs.\ ``figure'' vs.\ ``Fig.'': usually capitalized, make it uniform. And why not use the same graph as before?} The order of traversal of the graph is given by the value \textit{num} for each node of \GG. The \textit{lowlink} value is also displayed.

\begin{figure}[!h]
    \ctikzfig{lowlink}
    \ednote{sm}{couple of value $\rightarrow$ pair of values}
    \caption{A DFS was performed through \GG from $x_1$ and next to each node is represented the couple of value (\textit{num}, \textit{lowlink})\label{fig:exlowlink}}
\end{figure}

Now, backtracking edges can be highlighted w.r.t. the order of traversal. In fig. \ref{fig:exlowlink2}, they are represented as red dashed arrows.\\
Knowing the backtracking edges, all \textit{lowlink} values can be computed\footnote{In fact, they are refreshed during the DFS.}.\\
Let us take $x_8$ as an example: its \textit{lowlink} is equal to 2, which actually means that $x_6$\footnote{Because $\texttt{x6.num} = 2$} is its anchor -- or attachment node -- in their\edmargin{sm}{reference?} SCC, namely the green one. Indeed, $x_8$ is alone in its equivalence class\footnote{for the relation $\Rightarrow$},\edmargin{sm}{I don't understand the notion of equivalence, e.g. we have $x_8 \Rightarrow^* x_9 \Rightarrow^* x_8$.} and from all nodes linked by one of the backtracking edges of $x_8$, $x_6$ has the minimum value \textit{num}. Likewise, $x_3$ is its own attachment node in the green SCC since it is the first node visited when performing the DFS.

@@

\pagebreak
\section{Tarjan's algorithm}
\ednote{sm}{%
  Why present Tarjan's algorithm, which is not considered further here? Perhaps because it is the best known algorithm for computing SCCs? But then it should be compared with the set-based algorithm that this project is about?
}
Tarjan's algorithm is an efficient on-the-fly SCC computing algorithm \cite{tarjanMerz}. It basically perfoms a DFS while updating the \textit{num} and \textit{lowlink} values. All nodes are stored in a stack during the traversal until a backtracking edge is found. In this case, the \textit{lowlinks} are computed and all nodes are unstacked and saved in a SCC until a node verifying the equality between its \textit{num} and its \textit{lowlink} -- which has to occur -- is found. Then, the DFS goes on. The whole process is written in the following algorithm \ref{alg:tarjan}.\\
It can be shown that every node and edge are visited only once so the algorithm can achieve a linear complexity, \textit{i.e.} $\mathcal{O}(|\mathcal{V}|+|\mathcal{E}|)$.

\subsection{Description of Tarjan's algorithm}
\begin{algorithm}[H]\label{alg:tarjan}
    \SetAlgoLined
    \KwData{A graph \GG = (\VV, \EE)}
    \KwResult{A partition \texttt{SCCs} of \VV where each element of \texttt{SCCs} is a SCC of \GG}
    Initialize an empty stack \texttt{R}\;
    Initialize an empty set \texttt{SCCs}\;
    Let $\texttt{num} := 0$\;
    \ForAll{$v \in \mathcal{V}$}{
        \If{\texttt{v.num} is undefined}{
            SCC(v)\;
        }
    }
    \SetKwProg{Function}{function}{}{}
    \Function{SCC: $v \in \mathcal{V} \rightarrow \texttt{None}$}{
        \texttt{v.num = num}\;
        \texttt{v.lowlink = num}\;
        increment \texttt{num}\;
        Push $v$ in \texttt{R}\;
        \texttt{v.onStack = true}\;
        \ForAll{$w \in \text{POST}(v)$\quad\textcolor{red}{sm: POST undefined?}}{
            \If{\texttt{w.num} is undefined}{
                \texttt{v.lowlink} = $\min(\texttt{v.lowlink, w.lowlink})$ \quad\textcolor{red}{w.lowlink undefined?}\;
            }
            \ElseIf{\texttt{w.onStack}}{
                \texttt{v.lowlink} = $\min(\texttt{v.lowlink, w.num})$\;
            }
        }
        \If{\texttt{v.lowlink = v.num}}{
            Initialize an empty set \texttt{currentSCC}\;
            \Repeat{$\texttt{v.num} \neq \texttt{w.num}$\quad\textcolor{red}{sm: w.lowlink?}}{
                Let $w := \texttt{R.pop()}$\;
                \texttt{w.onStack = false}\;
                $\texttt{currentSCC} = \texttt{currentSCC} \cup \{w\}$\;
            }
            $\texttt{SCCs} = \texttt{SCCs} \cup \texttt{currentSCC}$\;
        }
    }
    \caption{Tarjan's algorithm}
\end{algorithm}
\pagebreak

\section{A sequential set-based algorithm}
\subsection{Formalisation}
\begin{definition}[SCC mapping]
    In the following algorithm, the SCCs are progressively tracked in a collection of disjoint\edmargin{sm}{We may have $\mathcal{S}(v) = \mathcal{S}(w)$ for $v \neq w$?!} sets through a map $\mathcal{S} : \mathcal{V} \longrightarrow \mathcal{P}(\mathcal{V})$, where $\mathcal{P}(\mathcal{V})$ is the powerset of \VV, s.t. the following invariant is maintained:
    \begin{equation}\label{invariant}
        \forall v, w \in \mathcal{V}, w\in \mathcal{S}(v) \Longleftrightarrow \mathcal{S}(v) = \mathcal{S}(w)
    \end{equation}
\end{definition}

@@

    \caption{Sequential set-based SCC algorithm}
\end{algorithm}
\ednote{sm}{%
  The algorithm should also explain how $\mathcal{S}$ is initialized.
}

@@

\subsection{Informal proof}
Note that this proof is said informal only because it is not formally automated.\edmargin{sm}{checked by a mechanized proof assistant?} Both logical and mathematical arguments developed below are absolutely relevant.

@@

\begin{proof}
    Let $x \in \mathcal{V}$ be the following node to be visited during the execution of the algorithm \ref{alg:seqsetbased}: $x$ is pushed in \texttt{R}. Let $y \in \textsc{Post}(x)$.\edmargin{sm}{The lemma assumes $y \in \texttt{R}$, is $y \in \textsc{Post}(x)$ an additional assumption? Why is it justified?} There are two cases:\\
    \begin{itemize}
        \item y has not been visited yet, \textit{i.e.} $y \notin \textsc{Visited}$. Thus, a DFS-like traversal is performed from $y$, so $y$ is pushed in \texttt{R} and $\mathcal{S}(y) = \{y\}$ because $y$ is alone in its equivalence class for the moment since it has not been visited yet.\\
        Therefore, $\mathcal{S}(x) \cap \mathcal{S}(y) = \varnothing$.\edmargin{sm}{I don't see an argument why $y \notin \mathcal{S}(x)$?}

        \item y has already been visited, \textit{i.e.} $y \in \textsc{Visited}$. Then, $y$ was already pushed in \texttt{R} before $x$. Let $(x_i)_{1\leq i \leq n}$ be the first nodes of the stack s.t. $x_0 = x$ and $x_n = y$.\edmargin{sm}{Should explain the order of the stack.}\\
        In order to avoid writing $\texttt{R} = [\ldots, y, \ldots, x]$, let us define $\widetilde{R}$ the stack containing the first $n$ nodes in \texttt{R}, s.t. $\widetilde{R} = [y, \ldots, x] = [x_n, \ldots, x_0]$.\\
        Let us consider the worst case, \textit{i.e.} when\edmargin{sm}{Why is this the worst case?}
        \begin{equation*}
            \forall~1 \leq i \leq n,~\mathcal{S}(x_i) = \{x_i\}
        \end{equation*}
        So, the while loop has to go down to $y$ because all partial SCCs are disjoint. As the length of the stack \texttt{R} is bounded by $|\mathcal{V}|$, the algorithm terminates.\\
        $x_0$ is first unstacked and both $\mathcal{S}(x_0)$ and $\mathcal{S}(\texttt{R.top()}) = \mathcal{S}(x_1)$ are then united. The current state of $\mathcal{S}$ and $\widetilde{R}$ is:\edmargin{sm}{Ambiguity between $\mathcal{S}$ as a mapping and as a set of equivalence classes.}
        \begin{equation*}
            \left\{
                \begin{array}{l}
                    \mathcal{S} = \{\{x_0, x_1\}, \{x_2\}, \ldots, \{x_n\}, \ldots\}\\
                    \widetilde{R} = [x_n, \ldots, x_1]\\
                \end{array}
            \right.
        \end{equation*}
        Then, $x_1$ is unstacked and $\mathcal{S}(x_1)$ and $\mathcal{S}(x_2)$ are then united, so that:
        \begin{equation*}
            \left\{
                    \begin{array}{l}
                        \mathcal{S} = \{\{x_0, x_1, x_2\}, \{x_3\}, \ldots, \{x_n\}, \ldots\}\\
                        \widetilde{R} = [x_n, \ldots, x_2]\\
                    \end{array}
                \right.
        \end{equation*}
        Finally (by induction), $\mathcal{S} = \{x_0, \ldots, x_n\}$ and $\widetilde{R} = [y]$, \textit{i.e.} $\mathcal{S}(x)=\mathcal{S}(y)$. It is important to notice that $x=x_0, x_1, \ldots, x_{n-1}$ are no longer in the stack, so this operation kept the invariant true.
    \end{itemize}
\end{proof}

@@

\begin{proof}
    The disjointness of all on-stack partial SCCs is given by lemma \ref{lemma:disjointness}. Nodes from $\textsc{Visited} \setminus \textsc{Explored}$ are in \texttt{R} because they are being processed. So, $\textsc{Live} \subseteq \texttt{R}$.\\
    By L.6-7 of algorithm \ref{alg:seqsetbased}, $\textsc{Visited} \subseteq \texttt{R}$.\\
    L.9-10 ensure that no explored node is pushed in \texttt{R}.\\
    L.24-25 keep the invariant by unstacking explored nodes from \texttt{R}, so $\texttt{R} \cap \textsc{Explored} = \varnothing$. Thus, $\texttt{R} = \textsc{Visited} \setminus \textsc{Explored} = \textsc{Live}$.
\end{proof}

\begin{corollary}\label{cor:cor1}
    \begin{equation*}
        \forall v \in \textsc{Live}, \exists!~r \in \texttt{R} \cap \mathcal{S}(v), \mathcal{S}(v) = \mathcal{S}(r)
    \end{equation*}
\end{corollary}
\begin{proof}
    Let $v \in \textsc{Live} = \displaystyle{\biguplus_{v \in \texttt{R}} \mathcal{S}(v)}$.\edmargin{sm}{Avoid using $v$ as a free variable and a bound variable in the same formula. Rather use $r$ for the bound variable, as in the lemma.} $v$ is in a unique partial SCC $\mathscr{S} := \mathcal{S}(v)$. Because of lemma \ref{lemma:disjointness}, there cannot exist $x \neq y \in \texttt{R}$ s.t. $\mathcal{S}(x) = \mathcal{S}(y) = \mathscr{S}$. Thus, there exists a unique $x \in \texttt{R}$ s.t. $\mathcal{S}(x) = \mathscr{S}$ (and $x\in \texttt{R} \cap \mathscr{S}$).
\end{proof}

@@

\begin{proof}
    Holds because of corollary \ref{cor:cor1}.\edmargin{sm}{In fact, just a weakening of Cor.1, why state it separately?}
\end{proof}
